\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-02-28 pea_paper_docs/tex/pea.tex

% use Unicode characters - try changing the option
% if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8x]{inputenc}
% clean citations
\usepackage{cite}
% hyperref makes references clicky. use \url{www.example.com}
% or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref}
% line numbers
\usepackage[right]{lineno}
% improves typesetting in LaTeX
% \usepackage{microtype}
% degree symbol
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{booktabs}
% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,
            labelsep=period,singlelinecheck=off]{caption}

% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
% \fancyheadoffset[L]{2.25in}
% \fancyfootoffset[L]{2.25in}

% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}

% define custom colors (this one is for figure captions)
\definecolor{Gray}{gray}{.25}

% this is required to include graphics
\usepackage{graphicx}
% set the figures directory
\graphicspath{{./figs/}}

% use if you want to put caption to the side of the figure - see example in text
\usepackage{sidecap}
% \usepackage[urlcolor  = blue]{hyperref}

@@

% new commands
\newcommand{\cel}{\emph{C.~elegans}}
\newcommand{\fog}{\emph{\mbox{fog-2}}}
\newcommand{\ecol}{\emph{E.~coli}}

\newcommand{\hobesity}{957}
\newcommand{\wobesity}{614}

\newcommand{\hlupus}{283}
\newcommand{\wlupus}{135}

\newcommand{\harthritis}{309}
\newcommand{\warthritis}{124}

@@

\author{David Angeles-Albores\textsuperscript{1,$\dagger{}$}
\and{}
Paul W. Sternberg\textsuperscript{1,*}
}

@@

]
% now start line numbers
\nolinenumbers{}

@@

The concept of a phenolog can also be useful when applied within a species.
In \cel{}, not all phenotypes are equally easy to study. Although genome-wide
measurements can help elucidate the genetic network underlying a phenotype,
devising screens to test which genes are functionally important can be difficult.
A common strategy to study phenotypes that are difficult to screen is to
select an easier-to-screen phenolog, and to test positive hits for the true
phenotype of interest afterwards~\cite{}. Currently, selection of screening
phenotypes is performed based on researcher experience.
By formalizing phenotype enrichment analysis as a tool with which to analyze
gene sets, researchers should be able to formally establish phenologs, which has
consequences for screen design.

@@

To facilitate understanding of large datasets, and to make discovery of
phenologs easier, we have completed an enrichment tool suite in WormBase
that allows users to rapidly perform phenotype, tissue and gene ontology
enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
using the same methodology for each one. They are located at
% TODO: include links.
We applied our tools towards the unbiased discovery of phenologues of
multigenic, complex diseases including
% Diseases to study:
systemic lupus erythematosus, obesity and obesity-related traits,  and
rheumatoid arthritis
by using genes associated with these
diseases via genome-wide association studies. We illustrate the utility of
the complete enrichment suite for finding new relationships in complex data by
analyzing a ciliary neuron transcriptome~\cite{Wang2015}. Finally, we show that
the dictionaries generated for these enrichment analyses can help elucidate the
contributions of specific tissues to specific phenotypes.

\section*{Methods}
\subsection*{Implementing the enrichment analyses}
All scripts were implemented in Python 3.5~\cite{Rossum2011}. We used
pandas~\cite{McKinney2011} and scipy~\cite{Oliphant2007} to write the
statistical testing framework. Matplotlib~\cite{Hunter2007} and Seaborn~\cite{Waskom}
libraries are used to generate all plots. Testing was performed using the
WormBase version XXXXX. The WormBase Enrichment suite can be installed using
pip via the command:

@@

A brief tutorial on how to use the command line tool can be found in the SI.
A brief tutorial on how to interact with the suite using python is available
as a Jupyter notebook~\cite{Perez2007} in the SI.

@@

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that
included terms with at
least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
(the total number of terms in the dictionary was 251, annotated by 9,169 genes
 for the version XXXXX);\@ and we generated a
dictionary that included terms with at least 50 annotating genes or more and
had a similarity threshold of 0.95 for GEA (the total number of terms in this
dictionary is 271, annotated by 14,636 genes for the version XXXX).
\@ Next, we benchmarked the dictionaries on the same gene
sets as TEA and obtained enrichment of all the expected
categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
Pauli2006, Portman2004, Fox2007, Smith2010}.
For example,
on a gene set enriched for embryonic muscle genes~\cite{Watson2008a},
the top two enriched
phenotype terms by q-value were `muscle system morphology variant' and `body
wall muscle thick filament variant'; the top two enriched GO terms were
`myofibril' and `striated muscle dense body'. For all the benchmarking
results, see supplementary information. Having generated and validated our
dictionaries, we proceeded to identify phenologs for several common human
diseases.

@@

The `short' phenotype was the result of the \emph{cat-4}, \emph{dpy-7},
\emph{rnt-1}, \emph{sem-4}, \emph{unc-116}, \emph{ocrl-1} and some genes in the
\emph{fat} family. Although these genes are bound by a common phenotype, any
genetic relationships between these genes are not immediately clear. Some genes,
like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
in development (including hypodermal development)~\cite{Desai1988,Ji2004}.
Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
the body of \cel{}. Yet others have known roles in neuron and muscle function,
such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
the `body length variant' phenotype. Body length in \cel{} can be controlled via
cell size or shape\cite{}; alternatively, cuticle development can alter body
shape~\cite{}; finally, muscles can alter the effective body length
due to their contraction state~\cite{}.

% TODO: SO.... ?

>>>>

% use Unicode characters - try changing the option
% if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8x]{inputenc}
% clean citations
\usepackage{cite}
% hyperref makes references clicky. use \url{www.example.com}
% or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref}
% line numbers
\usepackage[right]{lineno}
% improves typesetting in LaTeX
% \usepackage{microtype}
% degree symbol
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{booktabs}
% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}
% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,
            labelsep=period,singlelinecheck=off]{caption}
% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
% \fancyheadoffset[L]{2.25in}
% \fancyfootoffset[L]{2.25in}
% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}
% define custom colors (this one is for figure captions)
\definecolor{Gray}{gray}{.25}
% this is required to include graphics
\usepackage{graphicx}
% set the figures directory
\graphicspath{{./figs/}}
% use if you want to put caption to the side of the figure - see example in text
\usepackage{sidecap}
% \usepackage[urlcolor  = blue]{hyperref}

@@

% new commands
\newcommand{\cel}{\emph{C.~elegans}}
\newcommand{\fog}{\emph{\mbox{fog-2}}}
\newcommand{\ecol}{\emph{E.~coli}}
\newcommand{\hobesity}{957}
\newcommand{\wobesity}{614}
\newcommand{\hlupus}{283}
\newcommand{\wlupus}{135}
\newcommand{\harthritis}{309}
\newcommand{\warthritis}{124}

@@

\author{David Angeles-Albores\textsuperscript{1}
\and{}
Raymond YN Lee\textsuperscript{1}
\and{}
Juancarlos Chan\textsuperscript{1}
\and{}
Paul W Sternberg\textsuperscript{1,*}
}

@@

\textbf{1} Department of Biology and Biological Engineering,
and Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA\\
\textbf{*} Corresponding author. Contact:pws@caltech.edu

\section*{Abstract}
\bf{}
Genome-wide experiments have the capacity to generate massive amounts of unbiased
data about an organism. In order to interpret this data, dimensionality reduction
techniques are required. One approach is to annotate genes using controlled languages
and to test experimental datasets for term enrichment using probabilistic methods.
Although gene, phenotype and anatomy ontologies exist for \cel{}, no unified
software offers enrichment analyses of all the ontologies using the same methodology.
Here, we present the WormBase Enrichment Suite, which offers users the
ability to test all nematode ontologies simultaneously. We show that
the WormBase Enrichment Suite provides valuable insight into different
biological problems. Briefly, we show that phenotype enrichment analysis (PEA) can
help researchers identify disease phenologs, phenotypes that are homologous across species,
which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
analysis can also shed light on RNA-seq datasets by showing what
molecular functions are enriched, which phenotypes these functions are implicated
in and what tissues are overrepresented in the dataset. Finally, we explore the
phenotype-anatomy relationship, showing that a small subset of highly specific
tissues are disproportionately likely to cause an Egl phenotype, but inferring
tissue expression from an Egl phenotype is limited to the largest tissues.
]

% now start line numbers
\nolinenumbers{}

@@

The concept of a phenolog can also be useful when applied within a species.
In \cel{}, not all phenotypes are equally easy to study. Although genome-wide
measurements can help elucidate the genetic network underlying a phenotype,
devising screens to test which genes are functionally important can be difficult.
A common strategy to study phenotypes that are difficult to screen is to
select an easier-to-screen phenolog, and to test positive hits for the true
phenotype of interest afterwards. For example, candidate genes to extend the \cel{}
lifespan can be first screened for using heat shock survival
genes involved in \cel{} aging
sensitivity can be identified using stress assays~\cite{Kim2007a,Mehta2009}.
Currently, selection of screening phenotypes is performed based on researcher
experience. By formalizing phenotype enrichment analysis as a tool with which to
analyze gene sets, researchers should be able to formally establish phenologs,
which has consequences for screen design.
% eating defects in an ablated strain

@@

To facilitate understanding of large datasets, and to make discovery of
phenologs easier, we have completed an enrichment tool suite in WormBase
that allows users to rapidly perform phenotype, tissue and gene ontology
enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
using the same methodology for each one.
% They are located at
% % TODO: include links.
We applied our tools towards the unbiased discovery of phenologues of
multigenic, complex diseases including
% Diseases to study:
systemic lupus erythematosus, obesity and obesity-related traits,  and
rheumatoid arthritis
by using genes associated with these
diseases via genome-wide association studies. We illustrate the utility of
the complete enrichment suite for finding new relationships in complex data by
analyzing a ciliary neuron transcriptome~\cite{Wang2015}. Finally, we show that
the dictionaries generated for these enrichment analyses can help elucidate the
contributions of specific tissues to specific phenotypes.

\section*{Methods}
\subsection*{Implementing the enrichment analyses}
All scripts were implemented in Python 3.5~\cite{Rossum2011}. We used
pandas~\cite{McKinney2011} and scipy~\cite{Oliphant2007} to write the
statistical testing framework. Matplotlib~\cite{Hunter2007} and Seaborn~\cite{Waskom}
libraries are used to generate all plots. Testing was performed using the
WormBase version WS256. The WormBase Enrichment suite can be installed using
pip via the command:

@@

% A brief tutorial on how to use the command line tool can be found in the SI.
% A brief tutorial on how to interact with the suite using python is available
% as a Jupyter notebook~\cite{Perez2007} in the SI.

@@

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that
included terms with at
least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
(the total number of terms in the dictionary was 251, annotated by 9,169 genes
 for the version WS256);\@ and we generated a
dictionary that included terms with at least 50 annotating genes or more and
had a similarity threshold of 0.95 for GEA (the total number of terms in this
dictionary is 271, annotated by 14,636 genes for the version WS256).
\@ Next, we benchmarked the dictionaries on the same gene
sets as TEA and obtained enrichment of all the expected
categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
Pauli2006, Portman2004, Fox2007, Smith2010}.
For example,
on a gene set enriched for embryonic muscle genes~\cite{Watson2008a},
the top two enriched
phenotype terms by q-value were `muscle system morphology variant' and `body
wall muscle thick filament variant'; the top two enriched GO terms were
`myofibril' and `striated muscle dense body'. For all the benchmarking
results, see supplementary information. Having generated and validated our
dictionaries, we proceeded to identify phenologs for several common human
diseases.

@@

The `short' phenotype was the result of the \emph{cat-4}, \emph{dpy-7},
\emph{rnt-1}, \emph{sem-4}, \emph{unc-116}, \emph{ocrl-1} and some genes in the
\emph{fat} family. Although these genes are bound by a common phenotype, any
genetic relationships between these genes are not immediately clear. Some genes,
like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
in development (including hypodermal development)~\cite{Desai1988,Ji2004}.
Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
the body of \cel{}. Yet others have known roles in neuron and muscle function,
such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
the `body length variant' phenotype. Body length in \cel{} can be controlled via
cell size or shape\cite{Wang2002,Nakano2004}; alternatively, cuticle development can alter body
shape~\cite{Cox1980,Kramer1988}; finally, muscles can alter the effective body
length due to their contraction state~\cite{Lewis1980}.
